---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- read10xCounts(c(pbms5k= "/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/raw_feature_bc_matrix"))
sce
# the data is not 5k - might be 1k or so
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
sce <- sce[,colSums(counts(sce)) > 0]
sce
# assay(sce, "counts") = counts(sce)
``` 

- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
out <- DropletUtils::emptyDrops(counts(sce), lower = 100)
out
#could use ignore =0 arg for filtering counts = 0 (the previous step)
#test.ambient = T means to include cells that have lower RNA e.g. neutrophils 
```

**How do you read the output? Which droplets would you consider empty?**

> Answer: other rows should have >100UMI - the table is not showing all.
> 
>
>
> 

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}
is.cell <- out$FDR <= 0.01 # try different FDR(false discovery rate) e.g. 0.001 
sum(is.cell, na.rm=TRUE)
sce_filtered_emptydrops <- sce[ , which(is.cell), drop=F] # which -dropping NA and FDR <=0.01
sce_filtered_emptydrops
table(Sig=is.cell, limited=out$Limited)
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
# ncounts are not calculated in bioconductor - but Seurat does
library(scuttle)
sce <- scuttle::addPerCellQC(sce)
colData(sce)
sce$is.cell <- is.cell
length(sce)
colData(sce)

ggplot(as.data.frame(colData(sce))) +
  geom_violin(aes(x= is.cell, y= sum), scale = "width") +
  labs(x= "non empty droplets", y= "library size") +
  ylim(c(0,20000)) + # optional to trim the long tale of TRUE
  geom_point(aes(x= is.cell, y= sum)) + # optional
  geom_jitter(aes(x= is.cell, y= sum)) # optional

# FALSE: has been tested to be empty droplets but NA means the ssamples had less than 100 UMI so has not been tested
```

```{r}
is.mito <- grep("^MT-", rowData(sce)$Symbol); is.mito
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- read10xCounts(c(pbms_filtered = "/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/filtered_feature_bc_matrix"))
# to compare Cellranger filtered data and today's filtered
sce_filtered_cellranger
colData(sce_filtered_cellranger)

all_barcodes <- sce$Barcode
length(all_barcodes)
# total barcodes = 234600

compare_table <- tibble(
  emptydrops = all_barcodes %in% sce_filtered_emptydrops$Barcode   ,
  cellranger = all_barcodes %in% sce_filtered_cellranger$Barcode   ,
)
compare_table
table(compare_table)
# TRUE means good cells , overall good agreement (~10% good cells in both), if not in agreement, check the library size in the FALSE barcodes
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder::scDblFinder(sce_filtered_emptydrops) # many arg but can run without
colData(sce_scdblfinder)
table( sce_scdblfinder$scDblFinder.class  )
# my doublet number is 44 but others 30
hist(sce_scdblfinder$scDblFinder.score, breaks = 100)
# high score means they are more likely doublets
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> 
> 
> 

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

```{r}
sce_filtered_scdblfinder <- sce_scdblfinder[,sce_scdblfinder$scDblFinder.class == "singlet"]
sce_filtered_scdblfinder
table( sce_filtered_scdblfinder$scDblFinder.class)
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
# somehow sum column is gone so add
sce_filtered_scdblfinder <- scuttle::addPerCellQC(sce_filtered_scdblfinder)
colData(sce_filtered_scdblfinder)
table(sce_filtered_scdblfinder$scDblFinder.class)
colData(sce_scdblfinder)
# somehow sum column is gone so add
sce_scdblfinder <- scuttle::addPerCellQC(sce_scdblfinder)
lib_plot <- ggplot(as.data.frame(colData(sce_scdblfinder))) +
  geom_violin(aes(x= scDblFinder.class, y= sum), scale = "width") +
  labs(x= "singlet or doublet", y= "library size") 
lib_plot
```

```{r}
is.mito <- grep("^MT-", rowData(sce_scdblfinder)$Symbol)
sce_scdblfinder <- scuttle::addPerCellQC(sce_scdblfinder, subset =list("is.mito" =is.mito))
colData(sce_scdblfinder)
mt_plot <- ggplot(as.data.frame(colData(sce_scdblfinder))) +
  geom_violin(aes(x= scDblFinder.class, y= subsets_is.mito_percent), scale = "width") +
  labs(x= "singlet or doublet", y= "mt_percentage") 
mt_plot
```

```{r}
#sce_scdblfinder <- scuttle::addPerCellQC(sce_scdblfinder)
#pca_plot <- ggplot(as.data.frame(colData(sce_scdblfinder))) +
 # geom_violin(aes(x= scDblFinder.class, y= ), scale = "width") +
# labs(x= "singlet or doublet", y= "pca") 
#pca_plot
  
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}

set.seed(1)
sce_decontX <- celda::decontX(sce_filtered_scdblfinder,
                         z=NULL,
                         batch = NULL,
                         maxIter= 500 )
colData(sce_decontX)
#metadata(sce_decontX)
str(metadata(sce_decontX))
```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
#reducedDimNames(sce_decontX)
reducedDim(sce_decontX)
umap <- reducedDim(sce_decontX, "decontX_UMAP")
plotDimReduceCluster(x=sce_decontX$decontX_clusters, dim1 = umap[,1], dim2 = umap[,2])
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
colData(sce_decontX)
plotDecontXContamination(sce_decontX,
  batch = NULL,
  colorScale = c("blue", "green", "yellow", "orange", "red"),
  size = 1)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}
rownames(sce_decontX)
sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts")
rownames(sce_decontX) <- uniquifyFeatureNames(
  ID = rownames(sce_decontX),
  names = rowData(sce_decontX)[["Symbol"]])
# replace gene ID into gene symbol - if not matched - leave the gene ID
assayNames(sce_decontX)
plotDimReduceFeature(x=assay(sce_decontX, "logcounts"), 
                     features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1", "PPBP"),
                     dim1 = umap[,1], 
                     dim2 = umap[,2])
# the plots are before clean-up
```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
markers <- list(
  Tcell_Markers = c("CD3E", "CD3D"),
  Bcell_Markers = c("CD79A", "CD79B", "MS4A1"),
  Monocyte_Markers = c("S100A8", "S100A9", "LYZ"),
  NKcell_Markers = "GNLY"
)
cellTypeMappings <- list(
  Bcells = 2,
  Monocytes = c(1, 3),
  Tcells = 4,
  NKcells = c(5, 6)
)
# check cluster and cells since it changes every time. See umap clusters
plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("counts", "decontXcounts")
)

```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r}
plotDecontXMarkerExpression(
  
  
  
  
)
```

