---
title: "pathway_analysis"
author: "Inhye Park"
date: '2022-10-21'
output: html_document
---
This is the pathway analysis practical of the OBDS course 
- The first part of this practical is using gprofiler to perform ORA on output dataset of differentially expressed genes from DESeq2

# load packages 
```{r}
# load librarys

library(tidyverse)
library(gprofiler2)
```

# Read in our tables from this morning in a new Rmd file
```{r}
# read in files from this morning 

# 1. all results (this was filtered to remove non DE genes in our DESeq2 list so only contains genes expressed in our samples) -> call this table 'all_genes_df'  - this will be our background

all_genes_df <- read.csv("/project/obds/ipark/DESeq2_tutorial/CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")
head(all_genes_df)

#filter out the genes which padj is NA 

all_genes_df_filtered <- all_genes_df %>% 
    dplyr::filter(!is.na(padj))

# 2. all significant upregulated genes -> call this upreg_genes_df
upreg_genes_df <- read.csv("/project/obds/ipark/DESeq2_tutorial/CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_up.csv")
head(upreg_genes_df)

# 3. all significant down regulated genes -> call this downreg_genes_df 
downreg_genes_df <- read.csv("/project/obds/ipark/DESeq2_tutorial/CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_down.csv")
head(downreg_genes_df)

# Check the dimensions of each of your dataframes using dim()
dim(all_genes_df)
dim(upreg_genes_df)
dim(downreg_genes_df)
dim(all_genes_df_filtered)

# upreg_genes_df should be 915
# downreg_genes_df should be 1204
# all_genes_df_filtered should be 17729


```
Now we want a vector of gene ids to use for analysis 

```{r}
# from each of these dataframes get a vector of the ensembl ids 
class(all_genes_df_filtered)

all_gene_ids <- all_genes_df_filtered$ensenbl_gene_id
    
head(all_genes_df_filtered)

all_genes_df_filtered <- all_genes_df_filtered[,-1]
#all_genes_df_filtered <- all_genes_df_filtered[,2]
#head(all_genes_df_filtered)

#all_gene_ids <- all_genes_df_filtered$ , I need to remove one column named ensembl_gene_id and rename ensembl_gene_id
  
upreg_gene_ids <- upreg_genes_df$ensembl_gene_id
    
downreg_gene_ids <- downreg_genes_df$ensembl_gene_id
```

#Perform ORA using g:profiler 

1. Look at documentation - https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
```{r}
# use the gost() function - what paramters should we include/change 
# we want to test the ORA of upreg_gene_ids vs our all_gene_ids

# get help for function
?gost() 

# write query and assign to output variable named `gost_results_obj`

#gost_results_obj <- gost(query = upreg_gene_ids, 
                organism = "mmusculus", 
                ordered_query = FALSE, 
                multi_query = FALSE, 
                significant = TRUE, 
                exclude_iea = TRUE, 
                measure_underrepresentation = FALSE, 
                evcodes = FALSE, 
                user_threshold = 0.01, 
                correction_method = "g_SCS", 
                domain_scope = "annotated", 
                custom_bg = all_gene_ids, 
                numeric_ns = "", 
                sources = NULL, 
                as_short_link = FALSE)

gost_results_obj <- gost(query = upreg_gene_ids,
                organism = "mmusculus", 
                ordered_query = FALSE,
                multi_query = FALSE, 
                significant = FALSE, 
                exclude_iea = TRUE,
                measure_underrepresentation = FALSE, 
                evcodes = FALSE,
                user_threshold = 0.01, 
                correction_method = "g_SCS",
                domain_scope = "annotated", 
                custom_bg = all_gene_ids,
                numeric_ns = "", 
                sources = NULL, 
                as_short_link = FALSE)

# copy from the link above, changed exclude_iea, user_threshold, custom_bg, for correction_method g_SCS is recommended

# what are our outputs  
str(gost_results_obj)

```

```{r}
# plot our results using in built function in gprofiler -> which one is it? 
gostplot(gost_results_obj, capped = T, interactive = T)
```

```{r}
# what does as_short_link = FALSE do in gost() function? 
gost(query = upreg_gene_ids,
                organism = "mmusculus", 
                ordered_query = FALSE,
                multi_query = FALSE, 
                significant = FALSE, 
                exclude_iea = TRUE,
                measure_underrepresentation = FALSE, 
                evcodes = FALSE,
                user_threshold = 0.01, 
                correction_method = "g_SCS",
                domain_scope = "annotated", 
                custom_bg = all_gene_ids,
                numeric_ns = "", 
                sources = NULL, 
                as_short_link = TRUE)


```

Answer -> work on a web browser to get more info by as_short_link = TRUE

```{r}
# Filter results table for genesets with between 5-300 genes 



```


2. Try querying multiple gene lists at the same time (up and down regulated genes) using Multiple queries - see the documentation 

```{r}

```

Can you plot these querys together?

```{r}


```

Output the get_version_info() to save the versions of the databases that were used for the analysis!! 
```{r}
get_version_info(organism = "mmusculus") # if not specified, it will show human


```


```{r}
sessionInfo()

# save as Knit to keep the information
```

